---
title: "Hurricane Viz"
author: "Udit Ennam"
date: "April 30, 2018"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

* Hurricanes are one of the most deadliest natural disasters and highly unpredictable, with organizations across the world trying to predict the occurrences .

* 7 Tropical cyclone basins in the world:

North Atlantic

South Atlantic

North Indian

South Indian

East Pacific

South Pacific

West Pacific

To see whether there is a link between the number of hurricanes and the time of the year (here, we consider Month)

## Datasets

Used National Oceanic and Atmospheric Administration (NOAA) to extract datasets of 7 recognized hurricane basins in the world.

Datasets Parent URL : ftp://eclipse.ncdc.noaa.gov/pub/ibtracs/v03r03/wmo/csv/basin/

Major features we will look at:
Season, Basin, Name, Latitude, Longitude, Wind.WMO., Center, Month

```{r echo=FALSE}
# load packages
library(maps)
library(ggplot2)
library(dplyr)


# NOAA url
noaa = "ftp://eclipse.ncdc.noaa.gov/pub/ibtracs/v03r03/wmo/csv/basin/"

# 7 basins
basin = c("NA", "EP","NI","SA","SI","SP","WP")

# read files from IBTrACS
NA.basin = read.csv(paste(noaa, "Basin.", basin[1], ".ibtracs_wmo.v03r03.csv", sep = ""), skip = 1, stringsAsFactors = FALSE)
EP.basin = read.csv(paste(noaa, "Basin.", basin[2], ".ibtracs_wmo.v03r03.csv", sep = ""), skip = 1, stringsAsFactors = FALSE)
NI.basin = read.csv(paste(noaa, "Basin.", basin[3], ".ibtracs_wmo.v03r03.csv", sep = ""), skip = 1, stringsAsFactors = FALSE)
SA.basin = read.csv(paste(noaa, "Basin.", basin[4], ".ibtracs_wmo.v03r03.csv", sep = ""), skip = 1, stringsAsFactors = FALSE)
SI.basin = read.csv(paste(noaa, "Basin.", basin[5], ".ibtracs_wmo.v03r03.csv", sep = ""), skip = 1, stringsAsFactors = FALSE)
SP.basin = read.csv(paste(noaa, "Basin.", basin[6], ".ibtracs_wmo.v03r03.csv", sep = ""), skip = 1, stringsAsFactors = FALSE)
WP.basin = read.csv(paste(noaa, "Basin.", basin[7], ".ibtracs_wmo.v03r03.csv", sep = ""), skip = 1, stringsAsFactors = FALSE)



NA.basin = NA.basin[-1, ]
EP.basin = EP.basin[-1, ]
NI.basin = NI.basin[-1, ]
SA.basin = SA.basin[-1, ]
SI.basin = SI.basin[-1, ]
SP.basin = SP.basin[-1, ]
WP.basin = WP.basin[-1, ]

# column formatting
NA.basin$Season = as.numeric(NA.basin$Season)
NA.basin$Latitude = as.numeric(gsub("^ ", "", NA.basin$Latitude))
NA.basin$Longitude = as.numeric(gsub("^ ", "", NA.basin$Longitude))
NA.basin$Wind.WMO. = as.numeric(gsub("^ ", "", NA.basin$Wind.WMO.))
NA.basin$Wind.WMO.mph = NA.basin$Wind.WMO.*1.150779


EP.basin$Season = as.numeric(EP.basin$Season)
EP.basin$Latitude = as.numeric(gsub("^ ", "", EP.basin$Latitude))
EP.basin$Longitude = as.numeric(gsub("^ ", "", EP.basin$Longitude))
EP.basin$Wind.WMO. = as.numeric(gsub("^ ", "", EP.basin$Wind.WMO.))
EP.basin$Wind.WMO.mph = EP.basin$Wind.WMO.*1.150779


NI.basin$Season = as.numeric(NI.basin$Season)
NI.basin$Latitude = as.numeric(gsub("^ ", "", NI.basin$Latitude))
NI.basin$Longitude = as.numeric(gsub("^ ", "", NI.basin$Longitude))
NI.basin$Wind.WMO. = as.numeric(gsub("^ ", "", NI.basin$Wind.WMO.))
NI.basin$Wind.WMO.mph = NI.basin$Wind.WMO.*1.150779


SA.basin$Season = as.numeric(SA.basin$Season)
SA.basin$Latitude = as.numeric(gsub("^ ", "", SA.basin$Latitude))
SA.basin$Longitude = as.numeric(gsub("^ ", "", SA.basin$Longitude))
SA.basin$Wind.WMO. = as.numeric(gsub("^ ", "", SA.basin$Wind.WMO.))
SA.basin$Wind.WMO.mph = SA.basin$Wind.WMO.*1.150779

SI.basin$Season = as.numeric(SI.basin$Season)
SI.basin$Latitude = as.numeric(gsub("^ ", "", SI.basin$Latitude))
SI.basin$Longitude = as.numeric(gsub("^ ", "", SI.basin$Longitude))
SI.basin$Wind.WMO. = as.numeric(gsub("^ ", "", SI.basin$Wind.WMO.))
SI.basin$Wind.WMO.mph = SI.basin$Wind.WMO.*1.150779


SP.basin$Season = as.numeric(SP.basin$Season)
SP.basin$Latitude = as.numeric(gsub("^ ", "", SP.basin$Latitude))
SP.basin$Longitude = as.numeric(gsub("^ ", "", SP.basin$Longitude))
SP.basin$Wind.WMO. = as.numeric(gsub("^ ", "", SP.basin$Wind.WMO.))
SP.basin$Wind.WMO.mph = SP.basin$Wind.WMO.*1.150779


WP.basin$Season = as.numeric(WP.basin$Season)
WP.basin$Latitude = as.numeric(gsub("^ ", "", WP.basin$Latitude))
WP.basin$Longitude = as.numeric(gsub("^ ", "", WP.basin$Longitude))
WP.basin$Wind.WMO. = as.numeric(gsub("^ ", "", WP.basin$Wind.WMO.))
WP.basin$Wind.WMO.mph = WP.basin$Wind.WMO.*1.150779

# extract month for dataset NA.basin
time.date = strsplit(NA.basin$ISO_time, " ")
iso.date = unlist(lapply(time.date, function(x) x[1]))
iso.month = substr(iso.date, 6, 7)
NA.basin$Month = factor(iso.month, labels = c(month.name))

# extract month for dataset EP.basin
time.date = strsplit(EP.basin$ISO_time, " ")
iso.date = unlist(lapply(time.date, function(x) x[1]))
iso.month = substr(iso.date, 6, 7)
EP.basin$Month = factor(iso.month, labels = c(month.name)[-4])

# extract month for dataset NI.basin
time.date = strsplit(NI.basin$ISO_time, " ")
iso.date = unlist(lapply(time.date, function(x) x[1]))
iso.month = substr(iso.date, 6, 7)
NI.basin$Month = factor(iso.month, labels = c(month.name))

# extract month for dataset SA.basin
time.date = strsplit(SA.basin$ISO_time, " ")
iso.date = unlist(lapply(time.date, function(x) x[1]))
iso.month = substr(iso.date, 6, 7)
SA.basin$Month = factor(iso.month, labels = "March")

# extract month for dataset SI.basin
time.date = strsplit(SI.basin$ISO_time, " ")
iso.date = unlist(lapply(time.date, function(x) x[1]))
iso.month = substr(iso.date, 6, 7)
SI.basin$Month = factor(iso.month, labels = c(month.name))


# extract month for dataset SP.basin
time.date = strsplit(SP.basin$ISO_time, " ")
iso.date = unlist(lapply(time.date, function(x) x[1]))
iso.month = substr(iso.date, 6, 7)

SP.basin$Month = factor(iso.month, labels = c(month.name)[1:9])


# extract month for dataset WP.basin
time.date = strsplit(WP.basin$ISO_time, " ")
iso.date = unlist(lapply(time.date, function(x) x[1]))
iso.month = substr(iso.date, 6, 7)
WP.basin$Month = factor(iso.month, labels = c(month.name))

# joining the data frames
storms = rbind(NA.basin,EP.basin, NI.basin,SI.basin,WP.basin,SA.basin,SP.basin)
```

## Categorizing the hurricanes

We use Saffir-Simpson Scale to divide the hurricanes into 5 categories based on their wind speeds in miles per hour(mph)

```{r echo = FALSE}

brk<-c(-Inf,74,95,110,129,156,+Inf)
storms$cat<-cut(storms$Wind.WMO.mph,breaks=brk,include.lowest=T,include.highest=T,  labels=c(0,1,2,3,4,5))
```

Visualizing the number of hurricanes based on category

```{r echo=FALSE}

storms_max_cat <- storms %>% group_by(Serial_Num) %>% summarise(max=max(as.numeric(cat)))
strm_max_count <- storms_max_cat %>% group_by(max) %>% summarise(count = n())
strm_max_count$max <- strm_max_count$max - 1

saffir_simpson_colors <- c("#00faf4","#ffffcc","#ffe775","#ffc140","#ff8f20","#ff6060")
ggplot(data=strm_max_count, aes(factor(max), count)) + 
  geom_bar(stat="identity", fill = saffir_simpson_colors) +
  xlab("Category") + ylab("Number of hurricanes") + geom_text(stat='Identity',aes(label=`count`), vjust = -0.6, size = 3)

```


## Hurricane Visualization on a world Map across the 7 basins

Timeline being observed : 2000-2010

```{r echo=FALSE}
wm = maps::map("world", ".", exact = FALSE, plot = FALSE, fill = TRUE) %>% fortify()
```

```{r echo=FALSE}
substorms = subset(storms, Season %in% 2000:2010)
nop = which(substorms$Name == "NOT NAMED") # checking for which hurricanes to NOT be displayed
substorms = substorms[-nop, ]

# add and ID with name and season
substorms$ID = as.factor(paste(substorms$Name, substorms$Season, 
                               sep = "."))

# storm name as factor
substorms$Name = as.factor(substorms$Name)
```

```{r echo=FALSE}
map1 = ggplot(substorms, aes(x=Longitude, y=Latitude, group=ID)) + geom_polygon(data = maps::map("world", ".", exact = FALSE, plot = FALSE, fill = TRUE) %>% fortify(), aes(x=long,y=lat,group=group), fill="white", color = "gray", size = 0.2) + geom_path(data = substorms,aes(group=Serial_Num, color = cat), alpha = 0.4, size=0.3) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + xlab(NULL) + ylab(NULL) + facet_wrap(~Basin)
# maps
map1
```

Distribution of hurricanes across the 7 basins

```{r echo=FALSE}
storms_cat_basin <- storms %>% group_by(Serial_Num, Basin) %>% summarise(category = max(as.numeric(cat)))
storms_cat_basin_lst <- storms_cat_basin %>% group_by(Basin, category) %>% summarise(count = n())
ggplot(data = storms_cat_basin_lst, aes(Basin, count)) + geom_bar(stat= "identity", aes(fill = category)) + xlab('Basin') + 
ylab('Number of hurricanes [Category-wise]')

```


## Distribution of hurricanes around the United States of America

We look at the North Atlantic and North Pacific wind currents but across all the months of the year.

```{r echo=FALSE}
map2 = ggplot(substorms, aes(x=Longitude, y=Latitude, group=ID)) + geom_polygon(data = maps::map("world", ".", exact = FALSE, plot = FALSE, fill = TRUE) %>% fortify(), aes(x=long,y=lat,group=group), fill="white", color = "gray", size = 0.2) + geom_path(data = substorms,aes(group=ID, color = cat), alpha = 0.4, size=0.3) + xlim(-138, -20) + ylim(3, 55) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + xlab(NULL) + ylab(NULL)
map2
```

## Observation of occurrence of hurricanes over each month

```{r echo = FALSE}
map3 = ggplot(substorms, aes(x=Longitude, y=Latitude, group=ID)) + geom_polygon(data = maps::map("world", ".", exact = FALSE, plot = FALSE, fill = TRUE) %>% fortify(), aes(x=long,y=lat,group=group), fill="white", color = "gray", size = 0.2) + geom_path(data = substorms,aes(group=Serial_Num, color = cat), alpha = 0.4, size=0.3) +xlim(-138, -20) + ylim(3, 55) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + xlab(NULL) + ylab(NULL) + facet_wrap(~Month)
map3
```

## Future Work

* Combine more data from Wikipedia to check the affected areas and see if a particular area is affected in a specific timeline.

## Thank You!